---
title: "Moving out for a better career: Parental co-residence and later earnings"
output: html_notebook
params:
  pre_year: 2016
  post_year: 2017
  reform_year: 2015
  reform_month: 9
  bafog_threshold: 30000
---

```{r}
library(tidyverse)
library(modelsummary)

PATH_LISS_DATA <- "../../data/liss/"
PATH_LISS_TMP_DATA <- "../../tmp/liss/"

highest_educ <- read_rds(paste0(PATH_LISS_TMP_DATA, "liss_highest_educ.Rds")) %>% 
  filter(year >= 2018) %>% 
  arrange(nomem_encr, year) %>% 
  group_by(nomem_encr) %>% 
  summarize(
    highest_educ = dplyr::last(highest_educ)
  ) %>% 
  ungroup()

df <- read_rds(paste0(PATH_LISS_TMP_DATA, "liss_core.Rds")) %>%
  transmute(
    pid = nomem_encr,
    wave = zoo::as.Date(zoo::as.yearmon(as.character(wave), "%Y%m")),
    income_monthly = nettoink_f,
    positie = haven::zap_labels(positie),
    coresidence = case_when(
      positie == 5 ~ 1,                                # living with parents
      positie %in% c(1, 2, 3, 4, 6, 7) ~ 0,             # not living with parents
      positie == 9 ~ NA_real_,                          # missing/unknown
      .default = NA_real_
    ),
    birth_year = gebjaar,
    student = case_when(
      belbezig == 7 ~ 1,                               # enrolled as student
      .default = 0
    ),
    age = leeftijd
    # sex = as_factor(geslacht),
  )

```

## Naive model

In the naive model, I explain a person's earnings at the age of 30 with the number of years that they resided with their parents from age 18 to age 23.

where:

-   $i$ is the individual;

-   $\text{income}$ is the income earned by individual $i$ in log Euro at the age of 30;

-   $\text{coresidence}$ is the number of years the individual lived with their parents from age 18 to age 23;

-   $X$ is a vector of control variables, containing at least years of schooling and potential experience (Mincer equation)

$\hat{\beta}_1$ is likely biased, because parental co-residence during college years is not random. More conscientious individuals, for example, may prefer to live with their parents and will also earn more later, on average.

```{r}
later_income <- df %>% 
  filter(age == 30) %>% 
  select(pid, wave, income_monthly) %>% 
  group_by(pid) %>% 
  summarize(n_months = n(), income = log(sum(income_monthly, na.rm = TRUE) + 1)) %>% 
  ungroup() %>% 
  filter(n_months == 12) %>% 
  select(-n_months)


df_naive <- df %>% 
  filter(age >= 18 & age < 23) %>%
  group_by(pid) %>% 
  summarize(n_months = n(), coresidence = sum(coresidence, na.rm = TRUE) / 12, coresidence_sq = coresidence ** 2) %>% 
  ungroup() %>% 
  drop_na() %>% 
  full_join(later_income) %>% 
  drop_na() 

rm(later_income)

```

```{r}
naive_models <- list(
  "(1) log(income)" = lm(income ~ coresidence, data = df_naive),
  "(2) log(income)" = lm(income ~ coresidence + coresidence_sq, data = df_naive)
)

modelsummary(
  models = naive_models,
  coef_map = c(
    income = "log(income)",
    coresidence = "co-residence",
    coresidence_sq = "(co-residence)^2"
  ),
  gof_omit = "AIC|BIC|Log|RMSE",
  estimate  = "{estimate}{stars}",
  statistic = "({std.error})",
  stars = c('*' = .10, '**' = .05, '***' = .01)
)
```


```{r}
df %>% 
  filter(age >= 18 & age < 23) %>% 
  group_by(wave) %>% 
  summarize(
    coresidence = mean(coresidence, na.rm = TRUE),
    student = mean(student, na.rm = TRUE)
  ) %>% 
  ungroup() %>% 
  kableExtra::kable()
```

```{r}
# random_students <- df %>% 
#   filter(year(wave) == 2015 & student == 1) %>% 
#   sample_n(5) %>% 
#   pull(nomem_encr)
# 
# ggplot(df %>% filter(nomem_encr %in% random_students), aes(x = wave, y = coresidence)) +
#   geom_line() +
#   geom_point() +
#   facet_wrap(~nomem_encr)
```


```{r}
# ------------------------------------------------------------
# Identify respondents who are observed in all specified waves
# and belong to a given birth cohort.
#
# This is used to construct balanced panels around key
# educational transition periods.
# ------------------------------------------------------------
find_respondents_in_waves <- function(panel, birth_year, waves) {
  # TODO: Check continous appearance in waves
  return(
    panel %>% 
      # Restrict to a specific birth year
      filter(birth_year == .env$birth_year) %>% 
      # Keep only the waves of interest
      filter(wave %in% as.Date(waves)) %>% 
      # One record per person-wave
      distinct(nomem_encr, wave) %>% 
      # Count how many of the required waves are observed
      count(nomem_encr) %>% 
      # Keep respondents observed in all waves
      filter(n == length(waves)) %>% 
      # Return respondent identifiers
      pull(nomem_encr)
  )
}

START_PRE_REFORM <- "2014-09-01"
END_PRE_REFORM <- "2021-09-01"

# ------------------------------------------------------------
# Identify students from the pre-reform cohort:
# born in 1997 and observed at the start and end of the
# relevant study period.
# ------------------------------------------------------------
respondents_pre <- find_respondents_in_waves(
  panel,
  birth_year = 1997,
  waves = c(START_PRE_REFORM, END_PRE_REFORM)
)

START_POST_REFORM <- "2015-09-01"
END_POST_REFORM <- "2020-09-01"

# ------------------------------------------------------------
# Identify students from the post-reform cohort:
# born in 1998 and observed at the start and end of the
# relevant study period.
# ------------------------------------------------------------
respondents_post <- find_respondents_in_waves(
  panel,
  birth_year = 1998,
  waves = c(START_POST_REFORM, END_POST_REFORM)
)
```

