---
title: "Moving out for a better career: Parental co-residence and later earnings"
output: html_notebook
params:
  pre_year: 2014
  post_year: 2015
---

```{r}
library(tidyverse)
library(modelsummary)
library(lubridate)

PATH_LISS_DATA <- "../../data/liss/"
PATH_LISS_TMP_DATA <- "../../tmp/liss/"

highest_educ <- read_rds(paste0(PATH_LISS_TMP_DATA, "liss_highest_educ.Rds")) %>% 
  transmute(
    pid = nomem_encr,
    highest_educ = haven::zap_labels(highest_educ)
  ) %>% 
  group_by(pid) %>%
  summarize(
    wo  = as.numeric(any(highest_educ == 26, na.rm = TRUE)),                     # university degree
    hbo = as.numeric(any(highest_educ == 20, na.rm = TRUE) & !any(highest_educ == 26, na.rm = TRUE))
  ) %>% 
  ungroup()

df <- read_rds(paste0(PATH_LISS_TMP_DATA, "liss_core.Rds")) %>%
  transmute(
    pid = nomem_encr,
    wave = zoo::as.Date(zoo::as.yearmon(as.character(wave), "%Y%m")),
    income_monthly = nettoink_f,
    positie = haven::zap_labels(positie),
    coresidence = case_when(
      positie == 5 ~ 1,                                # living with parents
      positie %in% c(1, 2, 3, 4, 6, 7) ~ 0,             # not living with parents
      positie == 9 ~ NA_real_,                          # missing/unknown
      .default = NA_real_
    ),
    birth_year = gebjaar,
    student = case_when(
      belbezig == 7 ~ 1,                               # enrolled as student
      .default = 0
    ),
    age = leeftijd
    # sex = as_factor(geslacht),
  ) %>% 
  full_join(highest_educ)

rm(highest_educ)

```

## Naive model

In the naive model, I explain a person's earnings at the age of 30 with the number of years that they resided with their parents from age 18 to age 23.

where:

-   $i$ is the individual;

-   $\text{income}$ is the income earned by individual $i$ in log Euro at the age of 30;

-   $\text{coresidence}$ is the number of years the individual lived with their parents from age 18 to age 23;

-   $X$ is a vector of control variables, containing at least years of schooling and potential experience (Mincer equation)

$\hat{\beta}_1$ is likely biased, because parental co-residence during college years is not random. More conscientious individuals, for example, may prefer to live with their parents and will also earn more later, on average.

```{r}
later_income <- df %>% 
  filter(age == 30) %>% 
  select(pid, wave, income_monthly) %>% 
  group_by(pid) %>% 
  summarize(n_months = n(), income = log(sum(income_monthly, na.rm = TRUE) + 1)) %>% 
  ungroup() %>% 
  filter(n_months == 12) %>% 
  select(-n_months)

later_educ <- df %>% 
  select(pid, wo, hbo)

df_naive <- df %>% 
  filter(age >= 18 & age < 23) %>%
  group_by(pid) %>% 
  summarize(n_months = n(), coresidence = sum(coresidence, na.rm = TRUE) / 12, coresidence_sq = coresidence ** 2) %>% 
  ungroup() %>% 
  drop_na() %>% 
  full_join(later_income) %>% 
  full_join(later_educ) %>% 
  drop_na() 
  

rm(later_income)
rm(later_educ)

```

```{r}
naive_models <- list(
  "(1) log(income)" = lm(income ~ coresidence, data = df_naive),
  "(2) log(income)" = lm(income ~ coresidence + coresidence_sq, data = df_naive),
  "(3) log(income)" = lm(income ~ coresidence + coresidence_sq + hbo + wo, data = df_naive)
)

modelsummary(
  models = naive_models,
  coef_map = c(
    income = "log(income)",
    coresidence = "co-residence",
    coresidence_sq = "(co-residence)^2",
    hbo = "hbo (university of applied sciences)",
    wo = "wo (research university)"
  ),
  gof_omit = "AIC|BIC|Log|RMSE",
  estimate  = "{estimate}{stars}",
  statistic = "({std.error})",
  stars = c('*' = .10, '**' = .05, '***' = .01)
)
```



```{r}
find_respondents_in_waves <- function(df, birth_year, waves) {
  # TODO: Check continous appearance in waves
  return(
    df %>% 
      # Restrict to a specific birth year
      filter(birth_year == .env$birth_year) %>% 
      # Keep only the waves of interest
      filter(wave %in% as.Date(waves)) %>% 
      # One record per person-wave
      distinct(pid, wave) %>% 
      # Count how many of the required waves are observed
      count(pid) %>% 
      # Keep respondents observed in all waves
      filter(n == length(waves)) %>% 
      # Return respondent identifiers
      pull(pid)
  )
}

min_year <- params$pre_year - 5
max_year <- params$post_year + 5

cohorts <- purrr::map_dfr(
  seq(params$pre_year - 5, params$post_year + 5),
  function(year) {

    birth_year <- year - 18

    waves <- as.Date(c(
      paste0(year, "-09-01"),
      paste0(year + 5, "-09-01")
    ))

    pids <- find_respondents_in_waves(
      df,
      birth_year = birth_year,
      waves = waves
    )

    tibble(
      pid = pids,
      group = year
    )
  }
)

# TODO: Select based on student status instead of birth year
```

```{r}
df_did <- df %>% 
  left_join(cohorts) %>% 
  mutate(
    month_relative = interval(make_date(group, 9, 1), as_date(wave)) %/% months(1),
    quarter_relative = interval(make_date(group, 9, 1), as_date(wave)) %/% months(3),
    year_relative = interval(make_date(group, 9, 1), as_date(wave)) %/% months(12)
    # month_relative = case_when(
    #   group == params$pre_year ~ lubridate::interval(make_date(), lubridate::as_date(wave)) %/% months(1),
    #   group == params$post_year ~ lubridate::interval(as_date(params$post_window_start), lubridate::as_date(wave)) %/% months(1),
    #   .default = NA_real_
    # ),
    # quarter_relative = case_when(
    #   group == params$pre_year ~ lubridate::interval(as_date(params$pre_window_start), lubridate::as_date(wave)) %/% months(3),
    #   group == params$post_year ~ lubridate::interval(as_date(params$post_window_start), lubridate::as_date(wave)) %/% months(3),
    #   .default = NA_real_
    # ),
    # year_relative = case_when(
    #   group == params$pre_year ~ lubridate::interval(as_date(params$pre_window_start), lubridate::as_date(wave)) %/% months(12),
    #   group == params$post_year ~ lubridate::interval(as_date(params$post_window_start), lubridate::as_date(wave)) %/% months(12),
    #   .default = NA_real_
    # )
  ) %>% 
  # Keep only treated cohorts and the desired time window
  filter(!is.na(group)) %>% 
  filter(month_relative >= 0 & month_relative <= 60) %>% 
  select(-wave) %>% 
  transmute(
    pid,
    month_relative,
    quarter_relative,
    year_relative,
    group,
    student,
    coresidence,
    income_monthly,
    wo,
    hbo
  ) %>% 
  arrange(pid, month_relative)
```
```{r}
df_did_monthly <- df_did %>% 
  filter(wo == 1 | hbo == 1) %>% 
  group_by(month_relative, group) %>% 
  summarize(
    mean_core = mean(coresidence, na.rm = TRUE),
    sd_core   = sd(coresidence, na.rm = TRUE),
    n         = sum(!is.na(coresidence)),
    se_core   = sd_core / sqrt(n),
    ci_low    = mean_core - 1.96 * se_core,
    ci_high   = mean_core + 1.96 * se_core
  ) %>% 
  ungroup()

df_did_quarterly <- df_did %>% 
  filter(wo == 1 | hbo == 1) %>% 
  group_by(quarter_relative, group) %>% 
  summarize(
    mean_core = mean(coresidence, na.rm = TRUE),
    sd_core   = sd(coresidence, na.rm = TRUE),
    n         = sum(!is.na(coresidence)),
    se_core   = sd_core / sqrt(n),
    ci_low    = mean_core - 1.96 * se_core,
    ci_high   = mean_core + 1.96 * se_core
  ) %>% 
  ungroup()

df_did_yearly <- df_did %>% 
  filter(wo == 1 | hbo == 1) %>% 
  group_by(year_relative, group) %>% 
  summarize(
    mean_core = mean(coresidence, na.rm = TRUE),
    sd_core   = sd(coresidence, na.rm = TRUE),
    n         = sum(!is.na(coresidence)),
    se_core   = sd_core / sqrt(n),
    ci_low    = mean_core - 1.96 * se_core,
    ci_high   = mean_core + 1.96 * se_core
  ) %>% 
  ungroup()
```

```{r}

```


```{r}
# df_did_monthly %>% 
#   ggplot(aes(x = month_relative, y = mean_core, color = factor(group))) +
#   geom_line(linewidth = 1.25) +
#   # geom_point() +
#   scale_color_manual(
#     values = c("red", "orange"),
#     labels = c("Control: Pre-reform cohort", "Treatment: Post-reform cohort")
#   ) +
#   labs(
#     title = "Trends in co-residence (2016 Dutch study grant reform)",
#     x = "Time",
#     y = "Parental co-residence (%)",
#     color = "Cohort"
#   ) +
#   scale_y_continuous(limits = c(0, 1), labels = scales::percent) +
#   theme_minimal()

df_did_quarterly %>%
  filter(group %in% c(params$pre_year, params$post_year)) %>% 
  ggplot(aes(x = quarter_relative, y = mean_core, color = factor(group))) +
  geom_line(linewidth = 1.25) +
  # geom_point() +
  scale_color_manual(
    values = c("red", "darkgreen"),
    labels = c("Control: Pre-reform cohort", "Treatment: Post-reform cohort")
  ) +
  labs(
    title = "Trends in co-residence (2016 Dutch study grant reform)",
    x = "Time (Quarters)",
    y = "Parental co-residence (%)",
    color = "Cohort"
  ) +
  scale_y_continuous(limits = c(0, 1), labels = scales::percent) +
  theme_minimal()

df_did_yearly %>%
  filter(group %in% c(params$pre_year, params$post_year)) %>% 
  ggplot(aes(x = year_relative, y = mean_core, color = factor(group))) +
  geom_line(linewidth = 1.25) +
  # geom_point() +
  scale_color_manual(
    values = c("red", "darkgreen"),
    labels = c("Control: Pre-reform cohort", "Treatment: Post-reform cohort")
  ) +
  labs(
    title = "Trends in co-residence, comparing the 2014 to the 2015 cohort",
    x = "Time (years since start of study)",
    y = "Parental co-residence (%)",
    color = "Cohort"
  ) +
  scale_y_continuous(limits = c(0.5, 1), labels = scales::percent) +
  theme_minimal()
```
```{r}
df_did_monthly %>% 
  group_by(group) %>% 
  summarize(
    mean_core = mean(mean_core, na.rm = TRUE) * 5
  ) %>% 
  ungroup() %>% 
  ggplot(aes(x = group, y = mean_core)) +
  geom_line() +
  geom_point() +
  scale_x_continuous(breaks = seq(min(df_did_monthly$group), max(df_did_monthly$group), by = 1), minor_breaks = NULL) +
  scale_fill_manual(
    values = c(
      "2014" = "red",
      "2015" = "green"
    ),
    guide = "none"  # removes legend
  ) +
  annotate("rect",
           xmin = params$pre_year, xmax = params$post_year,
           ymin = -Inf, ymax = Inf,
           alpha = 0.2, fill = "lightblue") +
  ylim(0, 5) +
  labs(
    title = "Pre- and post-trends: Co-residence of study cohorts",
    x = "Cohort (Year start studying)",
    y = "Parental co-residence (years)"
  ) +
  theme_minimal() 
```


