---
title: "Coresidence-Earnings through BAföG"
author: "Oskar Veerhoek"
format: html
editor: visual
---

## R setup

```{r}
library(tidyverse)
library(modelsummary)
```

## Data

```{r}
df <- read_csv("../../tmp/soep/bafög_coresidence.csv", show_col_types = FALSE)

datasummary(
  formula = All(df) ~ Mean + SD + Min + Median + Max,
  data = df
)
```

## Naive model

$$
\log(\text{income})_{i} = \beta_0 + \beta_1 \,\text{coresidence}_{i} + \beta_2 X_{i} + \epsilon_{i}
$$

where:

-   $i$ is the individual\

-   $\text{income}$ is the income earned by individual $i$ in log Euro at age 30\

-   $\text{coresidence}$ is the number of years the individual lived with their parents from age 18 to 23\

-   $X$ is a vector of control variables

-   **Comment Robbert: Mensen die thuiswonen studeren sneller af**

    -   **Toevoeging André: Kijk naar time preferences**
    -   Traject / annual change i.p.v. level

-   **André: Gerritsen Paper**

$\hat{\beta}_1$ is likely biased, because parental co-residence during college years is not random.

## Simple difference-in-differences

$$
\text{coresidence}_{i} = \alpha + \beta_1 \text{eligible}_i + \beta_2 \text{post}_t 
+ \delta ( \text{eligible}_i \times \text{post}_t ) + \varepsilon_{i}
$$

```{r echo=FALSE, results='hide'}
df <- read_csv("../../tmp/soep/bafög_coresidence.csv") %>% 
  select(pid, syear, age, student, coresidence, grant, parental_income) %>% 
  filter(syear %in% c(2016, 2017)) %>% 
  filter(age >= 18 & age <= 23) %>% 
  filter(student == 1) %>% 
  select(-student) %>% 
  mutate(
    post = case_when(
      syear == 2016 ~ 0,
      syear == 2017 ~ 1,
      TRUE ~ NA_real_
    ),
    eligible = case_when(
      parental_income > log(30000) ~ 0,
      parental_income <= log(30000) ~ 1,
      TRUE ~ NA_real_
    )
  ) %>% 
  select(-syear, -age) %>% 
  arrange(pid, post) %>% 
  drop_na()
```

```{r}
library(fixest)

estimation <- feols(coresidence ~ eligible * post, data = df)

modelsummary(
  estimation,
  estimate  = "{estimate}{stars}",
  statistic = "({std.error})",
  stars = c('*' = .10, '**' = .05, '***' = .01)
)
```

## 
