---
title: "Moving out for a better career: Parental co-residence and later earnings"
output: html_notebook
---

## R setup

```{r}
library(tidyverse)
library(modelsummary)
library(fixest)
library(scales)

df <- read_csv("../../tmp/soep/bafög_coresidence.csv", show_col_types = FALSE)
```

## Data

### General summary statistics

```{r}
datasummary(
  formula = 
    (`Income (log Euro)` = income) + 
    (`Parental co-residence (dummy)` = coresidence) +
    (`BAföG grant received (Euro)` = grant) +
    (`Student status (dummy)` = student) +
    (`Age (years)` = age) +
    (`Parental income (log Euro)` = parental_income) +
    (`Years of schooling (years)` = educ)
    ~ N + PercentMissing + Mean + P25 + P50 + P75 + SD + Min + Max,
  data = df
)
```

### Eligibility for BAföG

```{r}
df %>% 
  mutate(parental_income_bin = round(parental_income)) %>% 
  filter(age >= 18 & age < 23 & student == 1) %>% 
  group_by(parental_income_bin) %>% 
  summarize(
    N = n(),
    grant = mean(grant, na.rm = TRUE)
  ) %>% 
  ungroup() 
```

## Naive model

$$
\log(\text{income})_{i} = \beta_0 + \beta_1 \,\text{coresidence}_{i} + \beta_2 X_{i} + \epsilon_{i}
$$

where:

-   $i$ is the individual;

-   $\text{income}$ is the income earned by individual $i$ in log Euro at the age of 30;

-   $\text{coresidence}$ is the number of years the individual lived with their parents from age 18 to age 23;

-   $X$ is a vector of control variables.

-   **Comment Robbert: Mensen die thuiswonen studeren sneller af**

    -   **Toevoeging André: Kijk naar time preferences**

    -   Traject / annual change i.p.v. level

-   **André: Gerritsen Paper**

$\hat{\beta}_1$ is likely biased, because parental co-residence during college years is not random. More conscientious individuals, for example, may prefer to live with their parents and will also earn more later, on average.

## Simple difference-in-differences

I will use a simple DiD analysis to estimate the effect of the 2016 BAföG reform on parental co-residence from the age of 18 to 23. The treatment group consists of students who are eligible for BAföG, because their parents earn below the income threshold. The control group consists of students who are not eligible for BAföG, because their parents earn above the income threshold.

$$
\text{coresidence}_{i} = \alpha + \beta_1 \text{eligible}_i + \beta_2 \text{post}_t 
+ \delta \left( \text{eligible}_i \times \text{post}_t \right) + \varepsilon_{i}
$$

**"stacked DID"**

**Robbert: Theorie specifieker, hoe shcerper je het verband hebt, hoe beter je**

where:

-   $i$ is the individual;

-   $\text{eligible}$ is whether the individual was eligible for BAfög

-   $\text{coresidence}$ indicates whether the individual was living with their parents;

-   $\text{post}_t$ is equal to 1 for observations in the period after the 2016 BAföG reform.

```{r echo=F, results='hide'}
df_simple_did <- df %>% 
  select(pid, syear, age, student, coresidence, grant, parental_income) %>% 
  filter(syear %in% c(2015, 2016)) %>% 
  filter(age >= 18 & age <= 23) %>% 
  filter(student == 1) %>% 
  select(-student) %>% 
  mutate(
    post = case_when(
      syear == 2015 ~ 0,
      syear == 2016 ~ 1,
      .default = NA
    ),
    eligible = case_when(
      parental_income > log(30000) ~ 0,
      parental_income <= log(30000) ~ 1,
      .default = NA
    )
  ) %>% 
  select(-syear, -age) %>% 
  arrange(pid, post) %>% 
  drop_na()

```

```{r}


estimation <- feols(coresidence ~ eligible * post, data = df_simple_did)

modelsummary(estimation,
             estimate  = "{estimate}{stars}",
            statistic = "({std.error})",
            stars = c('*' = .10, '**' = .05, '***' = .01),)
```

**Eligibility: fuzzy RD based on family income**

```{r}
df_simple_did %>% 
  group_by(eligible, post) %>% 
  summarize(mean_core = mean(coresidence, na.rm = TRUE)) %>% 
  ungroup() %>% 

  ggplot(aes(x = post, y = mean_core, group = eligible, color = factor(eligible))) +
  geom_line() +
  geom_point() +
  scale_color_manual(
    values = c("darkred", "darkblue"),
    labels = c("Not eligible", "Eligible")
  ) +
  scale_x_continuous(
    breaks = c(0, 1),
    labels = c("Pre-reform", "Post-reform")
  ) +
  labs(
    x = "",
    y = "Mean coresidence",
    color = ""
  ) +
  ylim(0, 1) +
  theme_minimal()
```

### Pre and post trends

```{r}
treatment_group <- df_simple_did %>% 
  filter(eligible == 1) %>% 
  pull(pid)

control_group <- df_simple_did %>% 
  filter(eligible == 0) %>% 
  pull(pid)

trends <- df %>% 
  select(pid, syear, age, student, coresidence, grant, parental_income) %>% 
  filter(syear >= (2016 - 5) & syear <= (2016 + 5)) %>% 
  filter(pid %in% c(treatment_group, control_group)) %>% 
  mutate(treatment = 
           case_when(
             pid %in% treatment_group ~ 1,
             pid %in% control_group ~ 0,
             .default = NA
           )) %>% 
  filter(!is.na(treatment)) %>% 
  group_by(syear, treatment) %>% 
  summarize(
    mean_core = mean(coresidence, na.rm = TRUE),
    sd_core   = sd(coresidence, na.rm = TRUE),
    n         = sum(!is.na(coresidence)),
    se_core   = sd_core / sqrt(n),
    ci_low    = mean_core - 1.96 * se_core,
    ci_high   = mean_core + 1.96 * se_core
    ) %>% 
  ungroup()

ggplot(trends, aes(x = syear, y = mean_core, color = factor(treatment))) +
  geom_line() +
  geom_point() +
  geom_errorbar(aes(ymin = ci_low, ymax = ci_high), width = 0.15) +
  annotate("rect",
           xmin = 2015, xmax = 2016,
           ymin = -Inf, ymax = Inf,
           alpha = 0.2, fill = "lightblue") +
  scale_color_manual(
    values = c("darkred", "darkblue"),
    labels = c("Control: Not eligible", "Treatment: Eligible")
  ) +
  scale_x_continuous(breaks = seq(min(df$syear), max(df$syear), by = 1), minor_breaks = NULL) +
  labs(
    title = "Pre- and post-reform trends in co-residence (2016 BAföG reform)",
    x = "Time",
    y = "Parental co-residence (%)",
    color = "Eligibility for BAföG in 2016"
  ) +
  scale_y_continuous(limits = c(0, 1), labels = percent) +
  theme_minimal()
  
```

Think of heterogeneity

Look at geo data
