---
title: "Estimating the Mincer equation with SOEP data"
output: html_notebook
---

# Estimating the Mincer earnings function with SOEP data

## The Mincer earnings function

The Mincer earnings function is given by:

$$
log\:w=as+bt-{ct}^2+other\:variables
$$

where:

-   $w$ is the worker's wage rate;

-   $s$ is the number of years in schooling;

-   $t$ gives the number of years of labor market experience;

-   $t^2$ is a quadratic on experience that captures the concavity of the age-earnings profile.

## Description of SOEP data

The German Socio-Economic Panel (SOEP) is a long-running longitudinal survey of households in Germany, conducted annually since 1984 by the German Institute for Economic Research (DIW Berlin).

In the context of estimating the Mincer earnings function, the SOEP is particularly valuable because it provides individual-level panel data on wages, education, and labor market experience, which are the core variables required in the Mincer framework.

## Constructing the data set

To construct the data set, we will use harmonized variables. DIW have helpfully harmonized several important variables across survey waves. We will use:

-   pgen: student enrollment,

-   pequiv: income, age, grant, number of years of education

-   pbrutto: parental co-residence

-   bioparen: parental income

### R initialization

```{r setup, include = FALSE}
knitr::opts_chunk$set(echo = FALSE)

library(tidyverse)
library(haven)
library(stargazer)
library(naniar)
library(modelsummary)

theme_set(theme_minimal())

SOEP_PATH <- "../../data/soep/SOEP-CORE.v40.1eu_Stata/Stata_DE/soepdata/"
```

### Loading data files

```{r load, include = FALSE}
pgen <- haven::read_dta(paste0(SOEP_PATH, "pgen.dta"))
pequiv <- haven::read_dta(paste0(SOEP_PATH, "pequiv.dta"))
bioparen <- haven::read_dta(paste0(SOEP_PATH, "bioparen.dta"))
pbrutto <- haven::read_dta(paste0(SOEP_PATH, "pbrutto.dta"))
bioedu <- haven::read_dta(paste0(SOEP_PATH, "bioedu.dta"))
ppathl <- haven::read_dta(paste0(SOEP_PATH, "ppathl.dta"))
pl <- haven::read_dta(paste0(SOEP_PATH, "pl.dta"))
```

### Extracting variables

```{r}
var1 <- pequiv %>% 
  transmute(
    pid,
    syear,
    
  )
```

## Naive estimation

First, we will

## Estimation with controls

## Estimation with parental co-residence
