---
title: "Estimating the Mincer equation with SOEP data"
output: html_notebook
---

# Estimating the Mincer earnings function with SOEP data

## The Mincer earnings function

The Mincer earnings function is given by:

$$
log\:w=as+bt-{ct}^2+other\:variables
$$

where:

-   $w$ is the worker's wage rate;

-   $s$ is the number of years in schooling;

-   $t$ gives the number of years of labor market experience;

-   $t^2$ is a quadratic on experience that captures the concavity of the age-earnings profile.

## Description of SOEP data

The German Socio-Economic Panel (SOEP) is a long-running longitudinal survey of households in Germany, conducted annually since 1984 by the German Institute for Economic Research (DIW Berlin).

In the context of estimating the Mincer earnings function, the SOEP is particularly valuable because it provides individual-level panel data on wages, education, and labor market experience, which are the core variables required in the Mincer framework.

## Constructing the data set

To construct the data set, we will use harmonized variables. DIW have helpfully harmonized several important variables across survey waves. We will use:

-   pgen: student enrollment,

-   pequiv: income, age, grant, number of years of education

-   pbrutto: parental co-residence

-   bioparen: parental income

### R initialization

```{r setup, include = FALSE}
knitr::opts_chunk$set(echo = FALSE)

library(tidyverse)
library(haven)
library(stargazer)
library(naniar)
library(modelsummary)

theme_set(theme_minimal())

SOEP_PATH <- "../../data/soep/SOEP-CORE.v40.1eu_Stata/Stata_DE/soepdata/"
```

### Loading data files

```{r load, include = FALSE}

files <- list(
  # i11110: individual labor income annual (Euro)
  pequiv = c("pid", "syear", "i11110", "d11109"),
  # sex: sex
  # gebjahr: year of birth
  # netto: current survey status
  # pop: sample membership
  # phrf: weighting factor
  ppathl = c("pid", "syear", "sex", "gebjahr", "netto", "pop", "phrf"),
  # pgtatzeit: actual work time per week
  # pgexpft: working experience fte
  # pgbilzeit: amount of education or traiing in years
  # pgfamstd: marital status
  pgen   = c("pid", "syear", "pgtatzeit", "pgexpft", "pgbilzeit", "pgfamstd", "pglabgro")
)

soep <- purrr::imap(
  files,
  ~ read_dta(
      file.path(SOEP_PATH, paste0(.y, ".dta")),
      col_select = all_of(.x)
    )
)

df_raw <- purrr::reduce(
  soep,
  dplyr::full_join,
  by = c("pid", "syear")
)

rm(soep)
rm(files)
```

### Description of loaded SOEP files

The following SOEP data files are used to construct the analysis dataset. Each file contains a different type of information about individuals or their family background.

**pgen.dta: `pid, syear, pgtatzeit, pgexpft, pgbilzeit, pgfamstd, pglabgro`**

The `pgen` file contains generated individual-level variables for each survey year. These include standardized labor market variables such as employment status, labor market experience indicators, and information on participation in education or training. The file is commonly used in empirical labor economics applications because it harmonizes key employment-related variables across waves.

**pequiv.dta: `i11110, d11109`**

The `pequiv` file provides equivalized income variables and several demographic characteristics. It includes measures of individual and household income, age, and years of education. These variables are central for estimating the Mincer equation because they allow us to construct wages, schooling measures, and demographic controls.

**ppathl.dta: `sex, gebjahr, netto, pop, phrf`**

The `ppathl` file contains information on individuals' life-course trajectories, particularly related to employment and labor market histories. It can be used to reconstruct employment spells and derive measures of labor market experience.

### Extracting variables

```{r}
df <- df_raw %>% 
  transmute(
    pid,
    syear,
    income_hourly = pglabgro,
    income_annual = i11110,
    educ_years = d11109,
    male = case_when(
      sex == 2 ~ 1,
      sex == 1 ~ 0,
      .default = NA
      ),
    birth_year = gebjahr,
    age = syear - birth_year,
    survey_status = netto,
    population_status = pop,
    weighting_factor = phrf,
    marital_status = pgfamstd,
    hours_worked_actual = pgtatzeit,
    work_experience_fte = pgexpft,
    education_training_years = pgbilzeit
  ) %>% 
  mutate(
    across(
      everything(),
      ~ replace (.x, .x %in% -9:-1, NA)
    )
  ) %>% 
  zap_label() %>% 
  zap_labels()
```

## Canonical cross-sectional Mincer model

$$
log\:w_i=\beta_0s_i+\beta_1t_i+c{t_i}^2+\epsilon_i
$$

```{r}
model1 <- df %>% 
  filter(syear == 2023) %>%
  filter(age >= 18 & age < 65) %>% 
  filter(income_hourly > 0) %>% 
  transmute(
    log_wage = log(income_hourly + 1),
    educ_years,
    experience = pmax(0, (syear - birth_year) - educ_years - 6),
    experience_sq = experience ** 2
  ) %>% 
  lm(log_wage ~ educ_years + experience + experience_sq, data = .) 

model1$coefficients[["educ_years"]]

modelsummary(
  models = list(
    "Canonical" = model1
  ),
  stars = c('*' = .1, '**' = .05, '***' = .01)
)
```

```{r}

df_reg <- df %>% 
    filter(age >= 18 & age < 65) %>% 
    filter(income_hourly > 0) %>% 
    transmute(
      syear,
      log_wage = log(income_hourly + 1),
      educ_years,
      experience = pmax(0, (syear - birth_year) - educ_years - 6),
      experience_sq = experience ** 2
    )

returns_to_education <- tibble()

for (year in min(df$syear):max(df$syear)) {
  educ_coefficient <- lm(
    formula = log_wage ~ educ_years + experience + experience_sq,
    data = filter(df_reg, syear == year)
  )$coefficients[["educ_years"]]
  returns_to_education <- bind_rows(returns_to_education, tibble(year = year, return_educ = educ_coefficient))
}
```

```{r}
ggplot(returns_to_education, aes(x = year, y = return_educ)) +
  geom_line()
```

## Estimation with controls

## Estimation with parental co-residence
