---
title: "Exploring SOEP Data"
output: html_notebook
---

# Exploring SOEP data

## Model

$$
log(income_{i,t+5})=\beta_0+\beta_1coresidence_it+\beta_2X_{it}+\epsilon_{it}
$$

$$\beta_1$$ is likely biased, because parental co-residence during college years is not random

## Variables of interest

In this Notebook, I explore different aspects of the SOEP household survey data set from DIW. My aim is to find longitudinal information on the following five variables:

1.  Age

2.  Income

3.  Study grant received, specifically BAföG (Bundesausbildungsförderungsgesetz)

4.  Parental co-residence

5.  Student enrollment

6.  Family income

7.  Level/duration of education

Let's start by loading the harmonized data sets:

```{r setup, include = FALSE}
knitr::opts_chunk$set(echo = FALSE)

library(tidyverse)
library(haven)
library(stargazer)
library(naniar)
library(modelsummary)

theme_set(theme_minimal())

SOEP_PATH <- "../../data/soep/SOEP-CORE.v40.1eu_Stata/Stata_DE/soepdata/"

pgen <- haven::read_dta(paste0(SOEP_PATH, "pgen.dta"))
pequiv <- haven::read_dta(paste0(SOEP_PATH, "pequiv.dta"))
bioparen <- haven::read_dta(paste0(SOEP_PATH, "bioparen.dta"))
pbrutto <- haven::read_dta(paste0(SOEP_PATH, "pbrutto.dta"))
bioedu <- haven::read_dta(paste0(SOEP_PATH, "bioedu.dta"))
ppathl <- haven::read_dta(paste0(SOEP_PATH, "ppathl.dta"))
```

## 1: Age

```{r}
age <- pequiv %>% 
  select(pid, syear, d11101) %>% 
  transmute(
    pid,
    syear, 
    age = case_when(
      d11101 < 0 ~ NA,
      .default = d11101
    )
  )
```

```{r}

age %>% 
  group_by(syear) %>% 
  summarize(
    n = n(),
    n_missing = sum(is.na(age)),
    age_min = min(age, na.rm = TRUE),
    age_max = max(age, na.rm = TRUE),
    age_median = median(age, na.rm = TRUE),
    age_mean = mean(age, na.rm = TRUE),
    age_sd = sd(age, na.rm = TRUE)
  ) %>% 
  knitr::kable()

```

## 2: Income

```{r}
income <- pequiv %>% 
  transmute(
    pid, 
    syear,
    income = case_when(
      i11110 < 0 ~ NA,
      .default = log(i11110 + 1)
    )
  )
```

```{r}
income %>% 
  group_by(syear) %>% 
  summarize(
    n = n(),
    n_missing = sum(is.na(income)),
    income_min = min(income, na.rm = TRUE),
    income_max = max(income, na.rm = TRUE),
    income_median = median(income, na.rm = TRUE),
    income_mean = mean(income, na.rm = TRUE),
    income_sd = sd(income, na.rm = TRUE)
  ) %>% 
  knitr::kable()
```

## 3: Study grant received

```{r}
grant <- pequiv %>% 
  transmute(
    pid,
    syear,
    grant = case_when(
      istuy < 0 ~ NA,
      .default = istuy
    )
  )
```

```{r}
grant %>% 
  filter(grant > 0) %>% 
  group_by(syear) %>% 
  summarize(grant = mean(grant, na.rm = TRUE)) %>% 
  ungroup() %>% 
  
  ggplot(aes(x = syear, y = grant)) +
  geom_line() +
  labs(
    title = "Grant of people who received more than 0 EUR",
    x = "Year",
    y = "Average grant received"
  ) +
  theme_minimal()
```

## 4. Parental co-residence

```{r}
coresidence <- pbrutto %>% 
  transmute(
    pid, 
    syear,
    coresidence = case_when(
      stell_h %in% c(20, 21) ~ 1,
      stell_h == 99 ~ NA,
      stell_h < 0 ~ NA,
      .default = 0
    )
  )
```

I should look into the jump before 2010. Is that because many older people were added then? Or perhaps because of some kind of harmonization issue? Consider household roster measure if this one is not reliable.

```{r}
coresidence %>% 
  group_by(syear) %>% 
  summarize(
    coresidence_perc = mean(coresidence, na.rm = TRUE) * 100
  ) %>% 
  ungroup() %>% 
  
  ggplot(aes(x = syear, y = coresidence_perc)) +
  geom_line() +
  labs(
    x = "Year",
    y = "Percentage parental coresidence"
  ) +
  theme_minimal()
```

## 5. Student enrolment

```{r}
student <- pgen %>% 
  transmute(
    pid,
    syear,
    student = case_when(
      pgstib == 11 ~ 1,
      pgstib < 0 ~ NA,
      .default = 0
    )
  )
```

Check whether this downward trend is because of changes to the underlying survey data.

```{r}
student %>% 
  group_by(syear) %>% 
  summarize(
    student_perc = mean(student, na.rm = TRUE) * 100
  ) %>% 
  ungroup() %>% 
  
  ggplot(aes(x = syear, y = student_perc)) +
  geom_line() +
  labs(
    x = "Year",
    y = "Percentage student enrolment"
  ) +
  theme_minimal()
```

## 6. Parental income

```{r}
parents <- bioparen %>% 
  transmute(
    pid,
    mnr1,
    fnr1
  )

parental_income <- pgen %>% 
  transmute(
    pid, syear
  ) %>% 
  left_join(parents, by = "pid") %>% 
  
  # mother income
  left_join(
    income %>% rename(mnr1 = pid, income_mother = income),
    by = c("mnr1", "syear")
  ) %>% 
  
  # father income
  left_join(
    income %>% rename(fnr1 = pid, income_father = income),
    by = c("fnr1", "syear")
  ) %>% 
  
  mutate(
    parental_income = coalesce(income_mother, 0) + coalesce(income_father, 0)
  )
  
parental_income <- parental_income %>% 
  transmute(
    pid, 
    syear,
    parental_income
  )
```

## 7. Number of years in education

```{r}
educ <- pequiv %>% 
  select(pid, syear, d11109) %>% 
  transmute(
    pid,
    syear,
    educ = case_when(
      d11109 < 0 ~ NA,
      .default = as.numeric(d11109)
    )
  )
```

## Combining variables

Tests:

1.  Check missing values for all variables

2.  Construct a cohort that I am interested in (pre-BAföG reform and post-BAföG reform)

3.  Association: Check the correlation between parental co-residence and later earnings)

4.  Check instrument strength (relation between reform and parental coresidence)

```{r}
df <- 
  full_join(age, coresidence) %>% 
  full_join(grant) %>% 
  full_join(income) %>% 
  full_join(student) %>%
  full_join(parental_income) %>% 
  full_join(educ) %>% 
  transmute(
    pid = as.numeric(pid),
    syear = as.numeric(syear),
    income = as.numeric(income), # Y: outcome variable
    coresidence = as.numeric(coresidence), # X: independent variable
    grant = as.numeric(grant), # IV
    student = as.numeric(student), 
    age = as.numeric(age),
    parental_income = as.numeric(parental_income),
    educ = as.numeric(educ)
  )
```

```{r}
any(duplicated(df))
any(duplicated(df %>% select(pid, syear)))
```

### Missing values

High missing values for the grant are to be expected, because the grant question is only asked to people who are students. High missings for age and student in final years are strange.

```{r}
df %>% 
  group_by(syear) %>% 
  summarize(
    missing_income = sum(is.na(income)) / n() * 100,
    missing_coresidence = sum(is.na(coresidence)) / n() * 100,
    missing_grant = sum(is.na(grant)) / n() * 100,
    missing_student = sum(is.na(student)) / n() * 100,
    missing_age = sum(is.na(age)) / n() * 100,
    missing_income_parent = sum(is.na(parental_income)) / n() * 100,
    missing_educ = sum(is.na(educ)) / n() * 100,
  ) %>% 
  ungroup() %>% 
  knitr::kable()
```

### Construct cohort

## **25th BAföG Amendment (25. BAföGÄndG) --- Key Features**

Most changes *took effect on **1 August 2016*** for the winter semester 2016/17 (the main BAföG period in Germany).

```{r}

students_wide <- df %>% 
  select(pid, syear, student) %>% 
  filter(syear %in% 2014:2018) %>% 
  pivot_wider(names_from = syear, values_from = student, names_prefix = "student_")

students_pre_reform <- students_wide %>% 
  filter(student_2014 == 0 & student_2015 == 1) %>% 
  pull(pid)

students_post_reform <- students_wide %>% 
  filter(student_2015 == 0 & student_2016 == 1) %>% 
  pull(pid)
```

Descriptive statistics of these students.

I'm only getting around 200 students who were students for the first time in a given academic year. I wonder if I need more for (a) the analysis and also (b) if the student variable is biased downwards.

```{r}
df %>% 
  filter(pid %in% students_pre_reform) %>% 
  filter(syear %in% 2015:2020) %>% 
  group_by(syear) %>% 
  summarize(n = n(), across(income:age, \(x) mean(x, na.rm = TRUE))) %>% 
  ungroup() %>% 
  knitr::kable()
```

Graphical visualization of the missings:

```{r}
naniar::gg_miss_var(df)
```

You can use BAföG receipt as a consistency check.

If BAföG recipients \>\> students classified\
→ your student definition is too strict.

```{r}
df %>% 
  filter(grant > 0 & syear == 2015) %>% 
  n_distinct()
```

### Association: Check the correlation between parental co-residence and later earnings)

```{r}
# Take age for simplicity
# Calculate parental coresidence in "college years" (18-22)

coresidence_proportion <- df %>% 
  filter(age >= 18 & age <= 22) %>% 
  group_by(pid) %>% 
  summarize(coresidence_proportion = mean(coresidence, na.rm = TRUE)) %>% 
  ungroup()
```

```{r}
hist(coresidence_proportion$coresidence_proportion)
```

```{r}
later_income <- df %>% 
  # Calcaulate income percentile to compare years
  group_by(syear) %>% 
  mutate(income_p = percent_rank(income)) %>% 
  ungroup() %>% 
  # Calculate later earnings as average over ages 28-32
  filter(age >= 28 & age <= 32) %>% 
  group_by(pid) %>% 
  summarize(later_income = mean(income_p, na.rm = TRUE)) %>% 
  ungroup()

hist(later_income$later_income)
```

Now combining later income and parental co-residence to calculate the correlation:

```{r}
association_df <- full_join(coresidence_proportion, later_income)

naniar::gg_miss_var(association_df)
```

```{r}
ggplot(association_df, aes(x = coresidence_proportion, y = later_income)) +
  geom_point() +
  geom_smooth()
```

# Simple DiD approach:

-   Outcome variable: parental co-residence

-   Treatment group: students with high parental income

-   Control group: students with low parental income

-   Time window: 2014, 2018

```{r}
df_did <- df %>% 
  filter(syear %in% c(2014, 2018)) %>% 
  filter(student == 1) 

```

```{r}
cor(df_did$grant, df_did$parental_income, use = "complete.obs")
```

```{r}
ggplot(df_did, aes(parental_income, grant)) +
  geom_smooth(method = "loess") +
  labs(
    x = "Parental Income (€)",
    y = "Average amount of BAföG Received"
  ) + 
  theme_minimal() +
  xlim(0, 100000)
```

```{r}

df_did <- df_did %>% 
  transmute(
    y = coresidence,
    treated = ifelse(parental_income < 25000, 1, 0), 
    post = ifelse(syear == 2018, 1, 0) 
  )
```

```{r}
did <- lm(y ~ treated * post, data = df_did)
stargazer(did, type = "text")
```

# Approach with all young people (and not just students)

## Naive model

$$
log(income_i)=\beta_0+\beta_1coresidence_i+\beta_2X_i+\epsilon_i
$$

where:

-   $log(income_i)$ is the income of person $i$ when they are 30 years old;
-   $coresidence_i$ is the number of years that person $i$ lived independently from 18 to 25;
-   $X_i$ are control variables;
-   $\epsilon_i$ is the error term.

## Stylized facts

### Fraction of young people living with their parents

```{r}
df %>% 
  group_by(age) %>% 
  summarize(
    n = n(),
    coresidence = mean(coresidence, na.rm = TRUE)
  ) %>% 
  ungroup() %>% 
  
  ggplot(aes(x = age, y = coresidence)) +
  geom_line() + 
  labs(
    x = "Age (years)",
    y = "Proportion of people living with their parents"
  ) +
  scale_y_continuous(labels=scales::percent)
```

```{r}
df %>% 
  filter(age >= 18 & age <= 25) %>% 
  group_by(syear) %>% 
  summarize(
    coresidence = mean(coresidence, na.rm = TRUE)
  ) %>% 
  ungroup() %>% 
  
  ggplot(aes(x = syear, y = coresidence)) +
  geom_line() + 
  labs(
    x = "Age (years)",
    title = "Proportion of people aged 18-25 living with their parents, by year"
  ) +
  scale_y_continuous(labels=scales::percent)
```

### Income distribution at 30 years

```{r}
df %>% 
  filter(age == 30) %>%
  group_by(syear) %>% 
  summarize(
    n = n(),
    income = mean(income, na.rm = TRUE),
    income = sd(income, na.rm = TRUE)
  )
```

### Differences in parental income

### Naive estimation

For the naive estimation, we construct the following variables:

-   Outcome variable: log wages at age 30

-   Treatment variable: parental co-residence, number of years a person lived with their parents between 18 and 30 (not including 30)

-   Covariates:

    -   Parental income: average log wages of parents when the person was between 18 and 30 years old

    -   Years of schooling: number of years of schooling, measured at 30 years old

```{r}
late_variables <- df %>% 
  filter(age == 30) %>% 
  select(pid, income, educ) %>% 
  transmute(
    pid,
    w = income,
    S = educ
  )

early_variables <- df %>% 
  filter(age >= 18 & age < 30) %>% 
  select(pid, coresidence, parental_income) %>% 
  group_by(pid) %>% 
  summarize(coresidence = mean(coresidence, na.rm = TRUE),
            parental_income = mean(parental_income, na.rm = TRUE)) %>% 
  ungroup()

df_naive <- full_join(late_variables, early_variables)

naniar::gg_miss_var(df_naive)
```

```{r}
df_naive <- drop_na(df_naive)

modelsummary::modelsummary(
  models = list(
    "Naive 1" = lm(income ~ coresidence, data = df_naive),
    "Naive 2" = lm(income ~ coresidence + educ, data = df_naive),
    "Naive 3" = lm(income ~ coresidence + educ + parental_income, data = df_naive)
  ),
  stars = c('*' = .1, '**' = .05, '***' = .01)
)
```

## Standard estimation Mincer earnings equation

### Necessary variables

-   Wage (hourly, monthly, or annual)

-   Years of schooling

-   Potential experience

### Data set

```{r}
df_mincer <- df %>%
  filter(age >= 18 & age < 65) %>% 
  transmute(
    lw = income,
    S = educ,
    Exp = pmax(0, age - S - 6),
    Exp2 = Exp ** 2
  )
```

```{r}
naniar::gg_miss_upset(df_mincer)
```

Removing missing values

```{r}
df_mincer <- drop_na(df_mincer)
```

Summary statistics

```{r}
datasummary_skim(df_mincer)
```

### Estimation of the basic Mincer equation

```{r}
model_mincer <- lm(lw ~ S + Exp + Exp2, data = df_mincer)
modelsummary(
  model_mincer,
  stars = c('*' = .1, '**' = .05, '***' = .01)
)
```
