---
title: "Population Density of Russia: The Siberian Effect"
output: html_notebook
---

## Motivation

A large share of Russia's territory lies in Siberia, a vast region known for its extreme climate and very low population density.  
Because Siberia covers such a large portion of the country, it strongly influences Russia's overall population density.

In this notebook we download public data from the **World Bank** to visualize how sparsely populated Russia is over time.

## Load packages

```{r setup}
library(WDI)
library(tidyverse)
```

## Download World Bank data

We use the World Bank indicator:

- **EN.POP.DNST** — Population density (people per square km of land area)

```{r data}
rus_density <- WDI(
  country = "RU",
  indicator = "EN.POP.DNST",
  start = 1960,
  end = 2023
) %>%
  as_tibble() %>%
  rename(pop_density = EN.POP.DNST)
```

## Plot population density over time

```{r plot}
ggplot(rus_density, aes(x = year, y = pop_density)) +
  geom_line(color = "steelblue", linewidth = 1.2) +
  geom_point(color = "darkred", size = 2) +
  labs(
    title = "Russia's Population Density Over Time",
    subtitle = "The vast and sparsely populated region of Siberia keeps national density very low",
    x = "Year",
    y = "People per square kilometer",
    caption = "Source: World Bank (WDI)"
  ) +
  theme_minimal()
```

## Interpretation

Russia's population density is extremely low compared to most countries.  
This is largely due to **Siberia**, which makes up the majority of Russia's land area but contains relatively few inhabitants.

The plot highlights how the country's population density has remained low and fairly stable over time.
